MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments

### 2021-10-11 main/MSstatsPTM_Main.tex
Added sample size and updated model section

Figure \ref{fig:data-structure} schematically illustrates a simplified version of the data structure resulting from a typical bottom-up experiment for quantitative analysis of PTMs, in which there are multiple layers of variation present. A PTM site is quantified with multiple spectral features, which vary in sequence (e.g., fully or partially cleaved peptides), ionization efficiency, charge states, etc. The number quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions. To perform adjustment with respect to protein abundance, features of unmodified peptides are used for the inference of underlying protein abundance. Typically, because of the enrichment step for PTMs, very few of those features are present in original LC-MS runs. For more accurate estimation of protein abundance, separate global proteomics data of unenriched samples are often acquired. If unmodified features are unavailable for any given modified feature, unmodified intensity adjustment cannot be performed. As different levels of variability are present in the data, the log-intensities of the features for modified and unmodified peptides are modeled separately using two linear mixed models.

@@

The proposed approach takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs. The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins. For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. In addition mixture may be included for features acquired via tandem mass tag (TMT) methods. The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance\cite{Choi:2014}. In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site\cite{Tukey:1977}. After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models\cite{Bolker2009} \cite{Faraway:2006}. Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.

@@

Detection of differentially modified PTM sites is performed through testing the null hypothesis of ‘no change’ against the alternative. The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance \ref{fig:data-structure}. Specifically, the adjusted difference is given by the difference in log-abundance of the PTM site, subtracted by the difference in log-abundance of the underlying protein, which is equivalent to the log of the ratio of PTM abundance difference to protein abundance difference. The estimate of the adjusted difference and the standard error (SE) of the estimate are obtained by combining the difference estimates and the associated SEs from both counterparts.

The test statistic for the hypothesis testing is the ratio of the estimate of the adjusted difference to its SE. To determine the statistical significance of the difference in terms of p-value, the test statistic is compared against the t distribution with degrees of freedom approximated by the Satterthwaite method\cite{satterthwaite:1946}. Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., $0.05$\cite{Benjamini:1995}. More details are provided in Supplementary Sec \ref{supp-sec:test}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Extension to TMT experiments}

The statistical modeling approaches discussed above can also be extended to Tandem Mass Tag (TMT) labeling methods. TMT experiments introduce an additional source of variation in the form of different mixtures. To account for mixture variance, first feature intensities are summarized again using Tukey's Median Polish and missing values are imputed, if desired, then a new linear mixed effects model with an added term for the mixture is included. The workflow and statisical model described generally follow the methods used for modeling proteins in $MSstatsTMT$\cite{Huang:2020}. These methods are repeated to quantify both modified and unmodified peptide abundance. Once both modified and unmodified peptides are modeled, the PTM model is adjusted for changes in global protein abundance using the same methods described previously.

For more information about the extension to Tandem Mass Tag experiments see Supplementary Sec. \ref{supp-sec:tmtmethod}.

@@

\todo{Devon: Update this section in main and supp}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}

@@

@@

>>>>

Figure \ref{fig:data-structure} schematically illustrates a simplified version of the data structure resulting from a typical bottom-up experiment for quantitative analysis of PTMs, in which there are multiple layers of variation present. A PTM site is quantified with multiple spectral features, which vary in sequence (e.g., fully or partially cleaved peptides), ionization efficiency, charge states, etc. The number quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions. To perform adjustment with respect to protein abundance, features of unmodified peptides are used for the inference of underlying protein abundance. Typically, because of the enrichment step for PTMs, very few of those features are present in original LC-MS runs. For more accurate estimation of protein abundance, separate global proteomics data of unenriched samples are recommended. If unmodified features are unavailable for any given modified feature, unmodified intensity adjustment cannot be performed. As different levels of variability are present in the data, the log-intensities of the features for modified and unmodified peptides are modeled separately using two linear mixed models. An overview of the steps of the proposed approach are illustrated in Figure \ref{fig:msstatsptm_design}.

@@

The proposed approach takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs. The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins. For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance\cite{Choi:2014}. This approach allows the method to be extended to cases where the experimental design in inbalanced, and where additional sources of variation are present. In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site\cite{Tukey:1977}. After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models\cite{Bolker2009} \cite{Faraway:2006}. Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.

@@

Detection of differentially modified PTM sites is performed through testing the null hypothesis of ‘no change’ against the alternative. The null hypothesis states that there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to protein abundance. Notationally, the hypothesis test is formulated as follows

\begin{align*}
H_{0}: \Delta = \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) - \left( \mu_{i} - \mu_{i'} \right) &= 0 \\
H_{a}: \Delta = \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) - \left( \mu_{i} - \mu_{i'} \right) &\neq 0
\end{align*}

Specifically, the adjusted difference is given by the difference in log-abundance of the PTM site, subtracted by the difference in log-abundance of the underlying protein, which is equivalent to the log of the ratio of PTM abundance difference to protein abundance difference. The estimate of the adjusted difference and the standard error (SE) of the estimate are obtained by combining the difference estimates and the associated SEs from both counterparts.

\[
SE(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}.
\]

Finally the degrees of freedom are given by

\[
\left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right)^2 \bigg/
\left( \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{\mathrm{df}(\gamma^{\ast})} + \frac{\hat{\sigma}_{\gamma}^{4}}{ \mathrm{df}(\gamma)} \right).
\]

The test statistic for the hypothesis testing is the ratio of the estimate of the adjusted difference to its SE. To determine the statistical significance of the difference in terms of p-value, the test statistic is compared against the t distribution with degrees of freedom approximated by the Satterthwaite method\cite{satterthwaite:1946}. Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., $0.05$\cite{Benjamini:1995}. More details on how the PTM model is adjusted are provided in Supplementary Sec \ref{supp-sec:test}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Extension to complex designs}

The statistical modeling approaches discussed above can also be extended to complex designs with additional sources of variation and inbalanced designs. One specific complex design the method can be applied to is Tandem Mass Tag (TMT) labeling methods. TMT experiments introduce an additional source of variation in the form of different mixtures. To account for mixture variance, first feature intensities are summarized again using Tukey's Median Polish and missing values are imputed, if desired, then a new linear mixed effects model with an added term for the mixture is included. The workflow and statisical model described generally follow the methods used for modeling proteins in $MSstatsTMT$\cite{Huang:2020}. These methods are repeated to quantify both modified and unmodified peptide abundance. Once both modified and unmodified peptides are modeled, the PTM model is adjusted for changes in global protein abundance using the same methods described previously.

For more information about the extension to complex designs and Tandem Mass Tag experiments see Supplementary Sec. \ref{supp-sec:complex_methods}.

@@

Here we analyze the sample size needed to achieve a desired statistical power. The proposed approach corrects for confounding between the modified and unmodified peptides by adjusting the modified peptide for changes in the unmodified. This adjustment comes at a cost of increased variation. This can be seen in the calculation for variance in Section \ref{sec:adjustment_summary}. When the variation is increased, the experiment requires a larger number of replicates to reach the same power. Thus the statistical power is dependent on the variance from both the modified peptide and unmodifed protein as well as the number of replicates in both runs.

We compared the statistical power in experiments with differing numbers of replicates, variance, and fold change for both the modified and unmodified runs. In terms of the number of replicates, we tested scenarios with equal replicates in both the modified and unmodified runs, as well as scenarios where the replicates differed between runs. For variance we used the biological experiments to determine what scenarios and values to test. In dataset 4 and 5 the variance of the PTM was higher than the global protein. In dataset 6 the variance of the PTM and Protein was generally the same due to not using a separate global profiling run. We mimicked these scenarios and analyzed the power of experiments when the PTM variance was higher than the protein and when they were equal. When the PTM and protein were the same we chose a variance of .15, whereas when the PTM was higher than the protein we chose a PTM variance of .2 and a protein variance of .1.

The results show that when the variance and replicates were equal, higher replicates predictably lead to higher power. In cases where the replicates were unbalanced, but the variance was still the same, it did not matter if there were more replicates in the modified or unmodified runs. As long as there were the same overall number of replicates the results were the same. In comparison, with differing variance and equal replicates, higher replicates still lead to higher power. When the replicates were unbalanced and the variance was higher in the PTM, there was more power when there were more replicates in the PTM than the protein. It is clearly more important to have high replicates for the PTM run than the unmodified protein, when the PTM variance is higher. In cases where the number of replicates has to be limited, it is better to weight them on the PTM side.

The results of the power and sample size analysis can be seen in Figure \ref{fig:power_sd_combo}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.8]{images/MSstatsPTM_design.png}
\caption{blah.}
\label{fig:msstatsptm_design}
\end{figure}

@@

\begin{figure}[h!]
\centering
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=.725\textwidth]{images/same_var_power}
	\caption{Here we can see the power of an experiment targeting PTMs with the same variance, .15, for the modified and unmodified peptides. Predictably when the replicates are high for both modified and unmodified peptides the power is much higher. Conversely at low replicates for each the power is much lower. The interesting part of this chart is when the replicates are different between runs. With equal variance, it does not matter if the PTM replicates or protein replicates are higher.}
 \end{subfigure}\vspace{-5mm}
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=.725\textwidth]{images/high_ptm_var_power}
	\caption{In this chart the variance for the PTM is higher than the unmodified protein. The PTM variance is .2, while the unmodified protein variance is .1. With equal replicates the results are the same as above, more replicates equals more power. When the replicates are not the same we can clearly see that having more replicates for the PTM leads to higher power.}
 \end{subfigure}
 \caption{Power analysis of experiments with differing variances.}
\label{fig:power_sd_combo}
\end{figure}
